Spontaneous breaking of multipole symmetries

### 2021-11-26 multipoleSSB/manuscript.tex
ready for submission
starting edits from rahul and ethan

\documentclass[twocolumn, longbibliography]{revtex4-2}

\usepackage{amsmath}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	linkcolor=blue,   
	urlcolor=blue,
}
\usepackage{graphicx}
\usepackage{tikz}

\newcommand{\note}[1]{\textcolor{red}{#1}}
\newcommand{\outline}[1]{\textcolor{Plum}{#1}}
\newcommand{\todo}[1]{\textcolor{orange}{#1}}
\newcommand{\charlie}[1]{\textcolor{TealBlue}{#1}}
\newcommand{\RN}[1]{\textcolor{TealBlue}{#1}}
\newcommand{\op}[1]{\mathcal{O}^{(#1)}}
\newcommand{\nn}{\nonumber\\}
\newcommand{\goesto}{\rightarrow}
\renewcommand{\max}{\text{max}}
\newcommand{\eff}{\text{eff}}
\newcommand{\cl}{\text{cl}}
\newcommand{\q}{\text{q}}
\newcommand{\half}{\frac{1}{2}}

@@

\title{Spontaneous breaking of multipole symmetries}
\author{Charles Stahl}
\author{Rahul Nandkishore}
\affiliation{Department of Physics and Center for Theory of Quantum Matter, University of Colorado Boulder, Boulder CO 80309, USA}

@@

This paper is organized as follows. In Section \ref{multipolegroup} we introduce the multipole group, how to build multipole-invariant field theories, and the generalized Mermin-Wagner argument for total breaking of the maximal multipole symmetry. In Section~\ref{sec:evade} we show how to evade the generalized Mermin-Wagner argument by considering phases in which multipole symmetries are partially broken.
We discuss generalized Imry-Ma arguments in the presence of quenched disorder in Section~\ref{sec:disord}. Finally, we consider an explicit lattice model that illustrates some of these ideas in Section~\ref{sec:example} before concluding with a discussion of open questions in \ref{sec:disc}.

@@

Finally, some of these multipole symmetries can be gauged to arrive at effective field theories for fracton phases. Of course, after gauging the theory will have gapless excitations due to the U(1) symmetry, but they can be Higgsed to arrive at a gapped phase. This is the sense in which the Haah group mentioned earlier corresponds to a field theory for Haah's code~\cite{BB}. See Ref.~\cite{Gromov2019} for the full story.

@@

@@

\section{Evading the generalized Mermin-Wagner argument} \label{sec:evade}

@@

The simplest case is when $a=1$ and $b=0$. In physical terms, we are taking a system that conserves total charge and total dipole moment, then condensing objects that carry a dipole moment but no total charge. We then recognize that these fundamental dipoles, which are of course the dipole charges of the monopole symmetry, are also the monopole charges of the dipole symmetry.  We should then expect symmetry breaking cannot occur for $d\le 2$ classically or $d\le 1$ quantum mechanically.

@@

In more generality, the order parameter for this pattern of symmetry breaking will be an object $\phi_{i_1\dots i_m}$, where $m=b+1$. Because $G$ is a maximal multipole group, we can still write the polynomials as in Eqn.~\ref{eqn:basis}. For our new order parameter, they act as
\begin{align}
\delta \phi_{i_1 \dots i_m} = \lambda_{I_c} \partial_{i_1} \dots \partial_{i_m} P_c^{I_c}, \label{eqn:partial}
\end{align}
so that $\phi_{i_1 \dots i_m}$ is invariant under $H=\mathcal{M}^b_\max$.

The kinetic term should take the form
\begin{align}
K[\phi(x)] &= (\partial_{i_{m+1}}\dots \partial_{i_n} \phi_{i_1 \dots i_m})^2,
\end{align}
which is invariant under shifts of the form of Eq.~\ref{eqn:partial}. The dispersion for the Goldstone modes is $K_k \sim k^{2(n-m)} = k^{2(a-b)}$. Following the correlation function calculation above, we find a classical critical dimension of $d = 2(a-b)$ and a quantum critical dimension of $d = a-b$. We recover the maximal generalized Mermin-Wagner argument if we let $b=-1$ denote the trivial group ($b=0$ denotes the monopole group).

\subsection{Mediated condensation} \label{sub:mediated}

We have seen that the generalized Mermin-Wagner argument provides critical dimensions for partial breaking that are lower than the critical dimensions for full symmetry breaking. In fact, we can use this observation to evade the generalized Mermin-Wagner theorem, completely breaking any multipole symmetry for $d>2$ classically, or $d>1$ quantum mechanically. This possibility has been pointed out in Refs.~\cite{Griffin2015Cascade, Griffin2015}, but we include a further interpretation in terms of what types of phase transitions are allowed in different dimensions. For simplicity, we will begin by considering the classical dipole-preserving theory with symmetry group $\mathcal{M}^2_\max$.

To be explicit, consider a monopole field $\phi(x)$ and a dipole field $\phi_i(x)$. Under the dipole symmetry they transform as 
\begin{align}
\phi(x) &\goesto \phi(x) + a_ix^i + b,\nn
\phi_i(x) &\goesto \phi_i(x) + a_i,
\end{align}
where we can more explicitly see that $\phi(x)$ transforms like a monopole field under the dipole part of the symmetry. The kinetic term will look like
\begin{align}
K[\phi,\phi_i] &= (\partial_i \partial_j \phi)^2 + (\partial_i \phi_j)^2,
\end{align}
so that $\phi$ has a $k^4$ dispersion and $\phi_i$ has a $k^2$ dispersion. We can also include an interaction term of the form $g(\partial_i \phi - \phi_i)^2$, which does not break the symmetry.  The $\phi_i$ field is well-defined whenever the dipole symmetry is broken. As long as $\phi_i$ is well-defined, integrating out fluctuations in $\phi_i$ endows the monopole field $\phi$ with a $k^2$ dispersion. The coupling $g$ is set by how far the theory sits from the dipole critical point.

With a $k^2$ dispersion for the monopole field, we now have a route to evade the (generalized) Mermin-Wagner argument and fully break the dipole symmetry in dimensions $d> 2$. First, partially break the symmetry so that $\phi_i$ is well defined, which is allowed for any $d>2$. Then, we can use the effective dispersion of the $\phi$ field to break the remaining part of the symmetry, also in any $d>2$. This procedure can be generalized, so that any multipole symmetry can be fully broken in $d>2$, provided we have enough symmetry-breaking order parameters. The discussion also translates to the quantum setting, so that the ultimate critical (space) dimension is always $d=1$.

The preceding procedure gives us insight into what the generalized Mermin-Wagner arguments were telling us. When we wrote down those arguments, we did not include the intermediate fields. Thus, the generalized Mermin-Wagner arguments tell us when a direct transition between two phases is or is not allowed. Consider again the dipole group $\mathcal{M}^2_\max$. The phase in which the symmetry is fully broken can exist in any dimension $d>2$ ($d>1$ for quantum systems), but there can only be direct phase transitions between this phase and the wholly symmetry unbroken phase in $d>4$ ($d>2$). In dimensions $2<d<4$ ($1<d<2$) the fully symmetry broken and the fully symmetry unbroken phase must be separated by an intermediate phase that breaks dipole but not monopole symmetry.

@@

As in the clean systems, we can have a sequence of phase transitions separated by intermediate phases in which the multipole symmetry is broken to a progressively smaller subgroup. With enough order parameter fields, any multipole symmetry can be fully broken in dimension $d>4$, regardless of which order parameter the disorder couples to. The same is true in quantum systems. Whether a certain phase transition is allowed or whether there needs to be an intermediate phase depends on the spatial dimension and the nature of the disorder.

@@

The most relevant Hamiltonian obeying these symmetries is  
\begin{align}
H_0  &= H_b + H_d + H_\text{int}\nn
H_b &= \sum_j \left[ \frac{U_b}{2}  n_j (n_j-1) - \mu_b \, n_j \right] \nn
&\qquad - \sum_{j,\mu,\nu} t_b \left[  b_j b^{\dag}_{j+\mu} b^{\dag}_{j+\nu} b_{j+\mu+\nu} + H.C. \right]\nn
H_d &= \sum_{j,\mu} \left[ \frac{U_d}{2} n'_{j,j+\mu} (n'_{j,j+\mu} - 1) -\mu \, n'_{j, j + \mu} \right]\nn
&\qquad  - \sum_{j,\mu,\nu} t_d \left[d^{\dag}_{j, j + \mu} d_{j + \nu, j+\mu + \nu} +H.C.\right]\nn
H_\text{int} &= \sum_{j,\mu} g\left[ b_j d_{j,j+\mu} b^\dag_{j+\mu} + H.C. \right],
\end{align}
where sums are taken over sites $j$ and lattice directions $\mu$ and $\nu$.  The operator $b^\dagger_j$ creates a boson on site $j$ while $d^\dagger_{j,j+\mu}$ creates an boson on the edge between $j$ and $j+\mu$. For simplicity, let us set 
\begin{align}
\frac{\mu_b}{ U_b} = \frac{\mu_d }{ U_d} = \frac{1}{2},
\end{align}
so there is a robust insulating phase at weak hopping $t_b, t_d$~\cite{Fisheretal}.

The Hamiltonian $H_b$ controls condensation of the site bosons. When $t_b$ is small, the site bosons are in number eigenstates while when $t_b$ is large the site bosons condense. Similarly, the edge bosons condense when $t_d$ is large. Note that all $d$ species of edge boson will condense equally, so that lattice rotation symmetry is not broken. The term $H_\text{int}$ is allowed by the symmetry because it simultaneously removes an edge boson and creates a dipole of site bosons. When the edge bosons are condensed, $H_\text{int}$ gives the site bosons an effective single particle hopping, and hence an effective $k^2$ dispersion, which would otherwise be disallowed by symmetry.

At sufficiently high dimension, this system will support three phases, which may be loosely described as the Mott insulator, the monopole superfluid where site bosons are condensed and the dipole superfluid where edge bosons (but not site bosons) are condensed. 
Note that the dipole superfluid does preserve lattice rotations.
In the monopole superfluid phase, the effective Hamiltonian contains bare creation and annihilation terms for the edge bosons, which automatically breaks number conservation of the edge bosons, and effectively induces a dipole superfluid as well. We can identify these phases with symmetry breaking patterns of the multipole group. In the trivial phase the group is unbroken. In the dipole superfluid, the group is broken to its monopole subgroup. The monopole superfluid then corresponds to the fully broken symmetry.

In the absence of disorder, neither symmetry breaking pattern can occur for $d\le 1$. There may still be interesting phases, such as a Luttinger liquid.
For $d>2$ dipole and monopole symmetry breaking may occur, so all three phases are intact. Furthermore, transitions between any two phases are allowed.
This leaves the intermediate case of $d=2$, in which all three phases can exist, but there can be no transitions between any symmetry-unbroken phase and the monopole superfluid. Instead, there must be an intermediate phase with dipole but not monopole superfluidity. Representative single-parameter phase diagrams are shown in Fig.~\ref{fig:phases}.

\begin{figure}
\centering

\begin{tikzpicture}[scale = .9]
\draw[<->] (0,0) node[left]{(a)}  -- (8,0) node[right]{$s$};
\draw (4,-.1) -- (4,.1);

\node [align=center] (A) at (2, -0.5) {unbroken};
\node [align=center] (C) at (6, -0.5) {fully\\broken};

\end{tikzpicture}
\begin{tikzpicture}[scale = .9]

\draw[<->] (0,0) node[left]{(b)}  -- (8,0) node[right]{$s$};
\draw (2.667,-.1) -- (2.667,.1);
\draw (5.333,-.1) -- (5.333,.1);

\node [align=center] (A) at (1.333, -0.5) {unbroken};
\node [align=center] (B) at (4, -0.5) {dipole\\broken};
\node [align=center] (C) at (6.667, -0.5) {fully\\broken};
\end{tikzpicture}

\caption{(a) For  $d>2$, there can be direct transitions between symmetry-unbroken and the fully-broken phase. Here $s$ is intended to represent a parameter along an arbitrary line in the phase diagram that crosses this transition. (b) At $d=2$, although the symmetry can be fully broken, the fully broken phase cannot touch the unbroken phase. Any line through the phase diagram that connects the two phases must pass though the dipole-broken phase. These phase diagrams are for translation invariant systems at zero temperature.}
\label{fig:phases}
\end{figure}

>>>>

\documentclass[pra,aps,twocolumn, amsfonts,amsmath,amssymb,nofootinbib,superscriptaddress]{revtex4}

\usepackage{amsmath}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	linkcolor=blue,   
	urlcolor=blue
}
\usepackage{graphicx}
\usepackage{tikz}

\newcommand{\note}[1]{\textcolor{red}{#1}}
\newcommand{\outline}[1]{\textcolor{Plum}{#1}}
\newcommand{\todo}[1]{\textcolor{orange}{#1}}
\newcommand{\charlie}[1]{\textcolor{Blue}{#1}}
\newcommand{\RN}[1]{\textcolor{TealBlue}{#1}}
\newcommand{\op}[1]{\mathcal{O}^{(#1)}}
\newcommand{\nn}{\nonumber\\}
\newcommand{\goesto}{\rightarrow}
\renewcommand{\max}{\text{max}}
\newcommand{\eff}{\text{eff}}
\newcommand{\cl}{\text{cl}}
\newcommand{\q}{\text{q}}
\newcommand{\half}{\frac{1}{2}}

@@

\title{Spontaneous breaking of multipole symmetries}
\author{Charles Stahl}
\affiliation{Department of Physics and Center for Theory of Quantum Matter, University of Colorado Boulder, Boulder CO 80309, USA}
\author{Ethan Lake}
\affiliation{Department of Physics, Massachusetts Institute of Technology, Cambridge, MA, 02139}
\author{Rahul Nandkishore}
\affiliation{Department of Physics and Center for Theory of Quantum Matter, University of Colorado Boulder, Boulder CO 80309, USA}

@@

This paper is organized as follows. In Section \ref{multipolegroup} we introduce the multipole group, how to build multipole-invariant field theories, and the generalized Mermin-Wagner argument for total breaking of the maximal multipole symmetry. In Section~\ref{sec:partial} we explore phases in which multipole symmetries are partially broken.
Then, we discuss generalized Imry-Ma arguments in the presence of quenched disorder in Section~\ref{sec:disord}. Finally, we consider an explicit lattice model that illustrates some of these ideas in Section~\ref{sec:example} before concluding with a discussion of open questions in \ref{sec:disc}.

@@

Finally, some of these multipole symmetries can be gauged to arrive at effective field theories for fracton phases. Of course, after gauging the theory will have gapless excitations due to the U(1) symmetry, but they can be Higgsed to arrive at a gapped phase. This is the sense in which the Haah group mentioned earlier corresponds to a field theory for Haah's code~\cite{BB}. See Ref.~\cite{BB, Gromov2019} for the full story.

@@

\charlie{Is it more useful to say $d_\text{cl} = d_\text{q} + z$ or $d_\text{cl} = 2 d_\text{q}$, or not try to make any general statement at all?}

@@

\section{Partial breaking of multipole symmetries} \label{sec:partial}

@@

The simplest case is when $a=1$ and $b=0$. In physical terms, we are taking a system that conserves total charge and total dipole moment, then condensing objects that carry a dipole moment but no total charge. \note{If $\phi(x)$ is charged under the monopole symmetry, then the operator $\partial_i \phi(x)$ is charged under the dipole part of the symmetry group. \charlie{Do we want correlation functions of $\partial_i\phi$ or $\exp(\partial_i\phi)$?} In a classical system at $T>0$ the correlation functions of this operator are
\begin{align}
\left\langle \partial_i\phi(x) \partial_j\phi(0) \right\rangle &\sim \delta_{ij} \int \frac{d^dk\, k^2}{k^4},
\end{align}
so that the critical dimension is $d=2$, as in the standard Mermin-Wagner theorem. }

%We then recognize that these fundamental dipoles, which are of course the dipole charges of the monopole symmetry, are also the monopole charges of the dipole symmetry.  We should then expect symmetry breaking cannot occur for $d\le 2$ classically or $d\le 1$ quantum mechanically.

@@

\note{In more generality, for this pattern of symmetry breaking there will be a charged operator $\partial_{i_1} \dots \partial{i_m} \phi$, where $m=b+1$. Because $G$ is a maximal multipole group, we can still write the polynomials as in Eqn.~\ref{eqn:basis}. As long as $\phi$ transforms as 
\begin{align}
\delta \phi = \lambda_{I_c} P_c^{I_c}, \label{eqn:partial}
\end{align}
then $\partial_{i_1} \dots \partial_{i_m} \phi$ is invariant under $H=\mathcal{M}^b_\max$. }

\note{Classical correlation functions for these operators take the form
\begin{align}
C &= \left\langle \partial_{i_1} \dots \partial_{i_m} \phi(x) \partial_{j_1} \dots \partial_{j_m} \phi(0) \right\rangle \nn
&\sim \int \frac{d^dk\, k^{2m}}{k^{2n}},
\end{align}
and diverge when $d \le 2(n-m) = 2(a-b)\equiv d_\text{cl}$. On the other hand, the $T=0$ correlation functions behave as
\begin{align}
C &= \left\langle \partial_{i_1} \dots \partial_{i_m} \phi(x) \partial_{j_1} \dots \partial_{j_m} \phi(0) \right\rangle \nn
&\sim \int \frac{d\omega\, d^dk\, k^{2m}}{\omega^2 + k^{2n}}\nn
&\sim \int \frac{d^dk\, k^{2m}}{k^n},
\end{align}
so the quantum critical dimension is $d_\text{q} = n - 2m = a - 2b - 1$. We recover the maximal generalized Mermin-Wagner argument if we let $b=-1$ denote the trivial group ($b=0$ denotes the monopole group). Furthermore, the critical dimensions again satisfy $d_\cl = d_\text{q} + z$. \charlie{Distinguish $z$ and $z_\max$?}
}

% The kinetic term should take the form
% \begin{align}
% K[\phi(x)] &= (\partial_{i_{m+1}}\dots \partial_{i_n} \phi_{i_1 \dots i_m})^2,
% \end{align}
% which is invariant under shifts of the form of Eq.~\ref{eqn:partial}. The dispersion for the Goldstone modes is $K_k \sim k^{2(n-m)} = k^{2(a-b)}$. Following the correlation function calculation above, we find a classical critical dimension of $d = 2(a-b)$ and a quantum critical dimension of $d = a-b$.

\subsection{No mediated condensation} \label{sub:mediated}

We have seen that the generalized Mermin-Wagner argument provides critical dimensions for partial breaking that are lower than the critical dimensions for full symmetry breaking. \note{Can we use this behavior to try to evade the Mermin-Wagner argument for the full symmetry group?}

%\remove{In fact, we can use this observation to evade the generalized Mermin-Wagner theorem, completely breaking any multipole symmetry for $d>2$ classically, or $d>1$ quantum mechanically}. This possibility has been pointed out in Refs.~\cite{Griffin2015Cascade, Griffin2015}, but we include a further interpretation in terms of what types of phase transitions are allowed in different dimensions. For simplicity, we will begin by considering the classical dipole-preserving theory with symmetry group $\mathcal{M}^2_\max$.

To be explicit, consider a monopole field $\phi(x)$ and a dipole field $\varphi_i(x)$. Under the dipole symmetry they transform as 
\begin{align}
\phi(x) &\goesto \phi(x) + a_ix^i + b,\nn
\varphi_i(x) &\goesto \varphi_i(x) + a_i.
\end{align}
The leading terms in the Lagrangian will be
\begin{align}
\mathcal{L} &= (\partial_t \phi)^2 + (\partial_t \varphi_i)^2 + (\partial_i \partial_j \phi)^2 + (\partial_i \varphi_j)^2,
\end{align}
so that $\phi$ has a $k^4$ dispersion and $\varphi_i$ has a $k^2$ dispersion. We can also include an interaction term of the form $g(\partial_i \phi - \varphi_i)^2$, which does not break the symmetry.  The $\varphi_i$ field is well-defined whenever the dipole symmetry is broken.

\note{We can then redefine the $\varphi_i$ field to $\tilde{\varphi}_i = \varphi_i - \partial_i \phi$, so that it is neutral under the symmetry. The interaction term becomes $g\tilde{\varphi}_i^2$, so that the $\tilde{\varphi}_i$ field becomes massive. The only remaining massless field is $\phi$, which remains charged under the full dipole group and keeps its $k^4$ dispersion. As a result, the critical dimension for breaking the full dipole group is not affected by the possibility of partially breaking.
}
\charlie{What else to say here? Compare to Griffin/Grosvenor?}

%As long as $\phi_i$ is well-defined, integrating out fluctuations in $\phi_i$ endows the monopole field $\phi$ with a $k^2$ dispersion. The coupling $g$ is set by how far the theory sits from the dipole critical point.

%With a $k^2$ dispersion for the monopole field, we now have a route to evade the (generalized) Mermin-Wagner argument and fully break the dipole symmetry in dimensions $d> 2$. First, partially break the symmetry so that $\phi_i$ is well defined, which is allowed for any $d>2$. Then, we can use the effective dispersion of the $\phi$ field to break the remaining part of the symmetry, also in any $d>2$. This procedure can be generalized, so that any multipole symmetry can be fully broken in $d>2$, provided we have enough symmetry-breaking order parameters. The discussion also translates to the quantum setting, so that the ultimate critical (space) dimension is always $d=1$.

%The preceding procedure gives us insight into what the generalized Mermin-Wagner arguments were telling us. When we wrote down those arguments, we did not include the intermediate fields. Thus, the generalized Mermin-Wagner arguments tell us when a direct transition between two phases is or is not allowed. Consider again the dipole group $\mathcal{M}^2_\max$. The phase in which the symmetry is fully broken can exist in any dimension $d>2$ ($d>1$ for quantum systems), but there can only be direct phase transitions between this phase and the wholly symmetry unbroken phase in $d>4$ ($d>2$). In dimensions $2<d<4$ ($1<d<2$) the fully symmetry broken and the fully symmetry unbroken phase must be separated by an intermediate phase that breaks dipole but not monopole symmetry.

@@

%\todo{No more mediated criticality}
%As in the clean systems, we can have a sequence of phase transitions separated by intermediate phases in which the multipole symmetry is broken to a progressively smaller subgroup. With enough order parameter fields, any multipole symmetry can be fully broken in dimension $d>4$, regardless of which order parameter the disorder couples to. The same is true in quantum systems. Whether a certain phase transition is allowed or whether there needs to be an intermediate phase depends on the spatial dimension and the nature of the disorder.

@@

The most relevant Hamiltonian obeying these symmetries is  
\begin{align}
H_0  &= H_b + H_d + H_\text{int}\nn
H_b &= \sum_j \left[ \frac{U_b}{2}  n_j (n_j-1) - \mu_b \, n_j \right] \nn
&\qquad - \sum_{j,\mu,\nu} t_b \left[  b_j b^{\dag}_{j+\mu} b^{\dag}_{j+\nu} b_{j+\mu+\nu} + H.C. \right]\nn
H_d &= \sum_{j,\mu} \left[ \frac{U_d}{2} n'_{j,j+\mu} (n'_{j,j+\mu} - 1) -\mu_d \, n'_{j, j + \mu} \right]\nn
&\qquad  - \sum_{j,\mu,\nu} t_d \left[d^{\dag}_{j, j + \mu} d_{j + \nu, j+\mu + \nu} +H.C.\right]\nn
H_\text{int} &= \sum_{j,\mu} g\left[ b_j d_{j,j+\mu} b^\dag_{j+\mu} + H.C. \right],
\end{align}
where sums are taken over sites $j$ and lattice directions $\mu$ and $\nu$.  The operator $b^\dagger_j$ creates a boson on site $j$ while $d^\dagger_{j,j+\mu}$ creates an boson on the edge between $j$ and $j+\mu$. For simplicity, let us set 
\begin{align}
\frac{\mu_b}{ U_b} = \frac{\mu_d }{ U_d} = \frac{1}{2},
\end{align}
so there is a robust insulating phase at weak hopping $t_b, t_d$~\cite{Fisheretal}.

The Hamiltonian $H_b$ controls condensation of the site bosons. When $t_b$ is small, the site bosons are in number eigenstates while when $t_b$ is large the site bosons condense. Similarly, the edge bosons condense when $t_d$ is large. Note that all $d$ species of edge boson will condense equally, so that lattice rotation symmetry is not broken. The term $H_\text{int}$ is allowed by the symmetry because it simultaneously removes an edge boson and creates a dipole of site bosons. When the edge bosons are condensed, $H_\text{int}$ gives the site bosons an effective single particle hopping, and hence an effective $k^2$ dispersion, which would otherwise be disallowed by symmetry. \todo{Yes to effective hopping, no to $k^2$ dispersion. Discuss SF/SC?}

\todo{Redo discussion of phases. Reframe in terms of symmetry-breaking patterns, not phases.}

\todo{Define multiple dipole-breaking patterns at the effective level. Relevance of cosine terms and stability to crystal phase?}
At sufficiently high dimension, this system will support three phases, which may be loosely described as the Mott insulator, the monopole superfluid where site bosons are condensed and the dipole superfluid where edge bosons (but not site bosons) are condensed. 
Note that the dipole superfluid does preserve lattice rotations.
In the monopole superfluid phase, the effective Hamiltonian contains bare creation and annihilation terms for the edge bosons, which automatically breaks number conservation of the edge bosons, and effectively induces a dipole superfluid as well. We can identify these phases with symmetry breaking patterns of the multipole group. In the trivial phase the group is unbroken. In the dipole superfluid, the group is broken to its monopole subgroup. The monopole superfluid then corresponds to the fully broken symmetry.

% In the absence of disorder, neither symmetry breaking pattern can occur for $d\le 1$. There may still be interesting phases, such as a Luttinger liquid.
% For $d>2$ dipole and monopole symmetry breaking may occur, so all three phases are intact. Furthermore, transitions between any two phases are allowed.
% This leaves the intermediate case of $d=2$, in which all three phases can exist, but there can be no transitions between any symmetry-unbroken phase and the monopole superfluid. Instead, there must be an intermediate phase with dipole but not monopole superfluidity. Representative single-parameter phase diagrams are shown in Fig.~\ref{fig:phases}.

% \begin{figure}
% \centering

% \begin{tikzpicture}[scale = .9]
% \draw[<->] (0,0) node[left]{(a)}  -- (8,0) node[right]{$s$};
% \draw (4,-.1) -- (4,.1);

% \node [align=center] (A) at (2, -0.5) {unbroken};
% \node [align=center] (C) at (6, -0.5) {fully\\broken};

% \end{tikzpicture}
% \begin{tikzpicture}[scale = .9]

% \draw[<->] (0,0) node[left]{(b)}  -- (8,0) node[right]{$s$};
% \draw (2.667,-.1) -- (2.667,.1);
% \draw (5.333,-.1) -- (5.333,.1);

% \node [align=center] (A) at (1.333, -0.5) {unbroken};
% \node [align=center] (B) at (4, -0.5) {dipole\\broken};
% \node [align=center] (C) at (6.667, -0.5) {fully\\broken};
% \end{tikzpicture}

% \caption{\todo{Remove figure} \todo{New figure with different dipole-breaking phases?}(a) For  $d>2$, there can be direct transitions between symmetry-unbroken and the fully-broken phase. Here $s$ is intended to represent a parameter along an arbitrary line in the phase diagram that crosses this transition. (b) At $d=2$, although the symmetry can be fully broken, the fully broken phase cannot touch the unbroken phase. Any line through the phase diagram that connects the two phases must pass though the dipole-broken phase. These phase diagrams are for translation invariant systems at zero temperature.}
% \label{fig:phases}
% \end{figure}
